### 2024-08-13 parts/correction.tex
starving the beast
correct proof of the lemma 15

Let us now prove that \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
This follows from the correct implication in the statement of \cref{karh:thm},
but we provide a self-contained proof using a refinement of the classical
combinatorial arguments for $\ZPoly$ \cite[Lemma 4.16]{CDTL23} and $\NPoly$
\cite[Lemma 5.37]{DOUE23}.

\begin{lemma}
    \label{n-poly-combinatorics:lem}
    Let $f$ be an \kl{$\Nat$-polyregular} function. 
    There exists a computable $\omega \in \Nat_{\geq 1}$
    such that for all $p \in \Nat$,
    for all $\alpha_0, \dots, \alpha_p \in \Sigma^*$,
    for all $u_1, \dots, u_p \in \Sigma^*$,
    there exists a polynomial $P \in \Rel[X_1, \dots, X_p]$
    whose \kl{maximal monomials} are \kl{non-negative},
    and such that for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \alpha_0 \prod_{i = 1}^p u_i^{\omega X_i} \alpha_i
        \right)
        = P(X_1, \dots, X_p) \quad .
    \end{equation*}
\end{lemma}
\begin{proof}
    \textbf{TODO: reprove}

    We leverage the characterization of \kl{$\Nat$-polyregular functions}
    given in \cref{nat-rel-poly:def}.
    Let $\omega$ be an idempotent power for the finite monoid $M$,
    and
    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$.
    Let $k \in \Nat$, and $\pi \colon M^k \topartial \Nat$ be such that
    $f = \pi^\dagger$. That is, 
    $f(w)$ is the sum over all factorizations of $w$
    into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
    of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.

    Let us define an equivalence relation $\equiv$ over words of
    $\Sigma^*$ as follows: two words $v_1, v_2$ are equivalent if they are
    equal when normalizing them with the rules $u_i^{\omega} u_i^{\omega} \to
    u_i^{\omega}$ for all $1 \leq i \leq p$. That is, we disregard
    repetitions of the factors $u_i^{\omega}$ in the words.
    We lift the equivalence relation $\equiv$ to $k$-tuples of
    words by pointwise application.

    Remark that if $(v_1, \dots, v_k) \equiv (w_1, \dots, w_k)$ then their
    production are the same, i.e., $\pi(\mu(v_1), \dots, \mu(v_k)) =
    \pi(\mu(w_1), \dots, \mu(w_k))$. This holds because $\mu(u_i^\omega) =
    \mu(u_i)^\omega$ is an idempotent element of the monoid $M$. In particular,
    $f(w)$ can be computed as the sum over equivalence classes for $\equiv$, of
    the value of $\pi$ on one representative, which belongs to $\Nat$.

    Therefore, it suffices to prove that the number of partitions of $w$ in a
    single equivalence class for $\equiv$ is a polynomial $P(X_1, \dots, X_p)$
    with \kl{non-negative} \kl{maximal monomials}.

    Notice that to produce all elements in a given equivalence class
    means adding the factors $u_i^\omega$ where $1 \leq i \leq p$
    back to the normalized version
    of the equivalence class. For all elements in an equivalence class,
    one has to add the same amount of $u_i^\omega$ for $1 \leq i \leq p$,
    and therefore the number of possible choices
    is either constant, or 
    is a product of binomials of the form $\binom{X - l}{s}$
    which almost has the desired shape.

    Indeed, the binomial 
    $\binom{X - l}{s}$ is equal to
    $(X-l) \cdots (X-l-s+1) / s!$.
    As a consequence,
    $\binom{s! X - l}{s} \in \Rel[X]$,
    and considering $\omega' \defined \omega \times s!$ (for some large enough $s$)
    allows us to conclude.
\end{proof}

@@

\begin{corollary}
    \label{decide-rat-poly-npoly:cor}
    Let $P \in \Rat[\vec{X}]$ be a polynomial with \emph{rational}
    coefficients and let $\alpha$ be the smallest number in $\Nat_{\geq 1}$
    such that $\alpha P \in \Rel[\vec{X}]$. Then, the following are equivalent:
    \begin{enumerate}
        \item $P$ is a \kl{$\Nat$-rational polynomial}.
        \item $\alpha P \in \CorrectPoly$,
            and for all $\vec{x} \in \Nat^{\vec{X}}$,
            $P(\vec{x})$ is a multiple of $\alpha$.
    \end{enumerate}
    In particular, the properties are decidable.
\end{corollary}
\begin{proof}
    If $P$ is a \kl{$\Nat$-rational polynomial}, then ...
    If $\alpha P \in \CorrectPoly$, then  it is an \kl{$\Nat$-rational polynomial}.
    But in particular, there exists a polyregular function $f$ that represents $\alpha P$
    with outputs in $\set{ a }$.
    Now, let us post-compose this function by the one that
    maps $a^{k \alpha + r}$ to $a^{k}$, which is a \kl{polyregular function}.
    We conclude that $P$ is \kl{represented} by an \kl{$\Nat$-polyregular function}.
\end{proof}

>>>>

\AP Let us now prove that \kl{$\Nat$-rational polynomials} are in
$\CorrectPoly$. This follows from the correct implication in the statement of
\cref{karh:thm}, but we provide a self-contained proof using a refinement of
the classical combinatorial arguments for $\ZPoly$ \cite[Lemma 4.16]{CDTL23}
and $\NPoly$ \cite[Lemma 5.37]{DOUE23}. We believe that the formulation of such
pumping arguments can be made clearer through some extra notations, with the
added property of highlighting the central role of \kl{commutative} polyregular
functions. Let us define that a function $q$ is a \intro{(noncommutative output)
polynomial function} from $\Nat^p$ to $\Sigma^*$ whenever there exists words
$\alpha_0, \dots, \alpha_p \in \Sigma^*$, and words $u_1, \dots, u_p \in
\Sigma^*$, such that $q(X_1, \dots, X_p) = \alpha_0 \prod_{i = 1}^p u_i^{X_i}
\alpha_i$. That is, $q$ is syntactically defined by a non-commutative
polynomial function over the semi-ring $\Sigma^*$. Almost by definition,
\kl{polynomial functions} are \kl{commutative} \kl{polyregular functions}.

\AP There is still a minor detail that remains to be smoothened before proving
our combinatorial \cref{n-poly-combinatorics:lem}, which is that some
polyregular functions can exhibit a limited form of division. For instance, the
function computing the sum of the lengths of the suffixes of a word $w$
actually computes $\card{w} \times (\card{w} - 1) / 2$. This will be at odds
with our restriction to polynomials with coefficients in $\Rel$. However, we
argue that up to a suitable change of basis for the ring of polynomials, all
coefficients will actually be in $\Nat$. This relatively simple trick works by
considering as ``basic polynomials" the \intro{binomial polynomials} of the
form $\binom{X - \ell}{k}$, where $\ell$ and $k$ are natural numbers. This is
why we introduce the notion of \intro{natural binomial polynomial} as
$\Nat$-linear combinations of products of \kl{binomial polynomials}.

\begin{lemma}
    \label{n-poly-combinatorics:lem}
    Let $f$ be an \kl{$\Nat$-polyregular} function. 
    There exists a computable $\omega \in \Nat_{\geq 1}$
    such that for all \kl{polynomial functions}
    $q \colon \Nat^p \to \Sigma^*$,
    there exists a computable \kl{natural binomial polynomial} $P$
    such that:
    \begin{equation*}
        f \circ q(\omega X_1, \dots, \omega X_p)
        =
        P
        \quad 
        \text{ over } (\Nat_{\geq 1})^p
    \end{equation*}
\end{lemma}
\begin{proof}
    Let us fix a finite monoid $M$, a morphism $\mu \colon \Sigma^* \to M$, and 
    let $\omega$ be an idempotent power for the finite monoid $M$,
    i.e., a number such that for all $x \in M$,
    $x^{\omega+1} = x^\omega$.
    We will prove by induction on $k \in \Nat_{\geq 1}$
    that for all
    $\pi \colon M^k \to \Nat$,
    the lemma holds for the function 
    $f = \pi^\dagger$, i.e., such that
    $f(w)$ is the sum over all factorizations of $w$
    into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
    of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.
    Since every \kl{$\Nat$-polyregular function}
    is obtained via some choice of $M, \mu, \pi$ (\cref{nat-rel-poly:def})
    we derive the desired result.

    In all cases, we will use a \kl{polynomial function} $q \colon \Nat^p \to
    \Sigma^*$ defined by $q(X_1, \dots, X_p) \defined \alpha_0 \prod_{i = 1}^p
    (u_i^{\omega \times X_i} \alpha_i)$ using some words $\alpha_i$ ($0 \leq i
    \leq p$) and $u_i$ ($1 \leq i \leq p)$ in $\Sigma^*$. That is, we will
    implicitly multiply all variables by $\omega$ in order to make the
    equations more readable. Note that when using an induction hypothesis, we
    will therefore have to check that the polynomial $q$ will have this
    specific form.

    \textbf{Base case: $k = 1$.}
    Let us show that $f \circ q$ is
    actually a constant (positive) function (and in particular a \kl{natural binomial polynomial}), as for
    all $n_1, \dots, n_p \geq 1$,
    \begin{align*}
        f(q(n_1, \dots, n_p))
        &\eqdef \pi\left(\mu(q(n_1, \dots, n_p))\right) \\
        &\eqdef \pi\left(\mu(\alpha_0 \prod_{i = 1}^p u_i^{\omega n_i} \alpha_i)\right)
        \\
        &= \pi\left(\mu(\alpha_0) \prod_{i = 1}^p \mu(u_i)^{\omega n_i} \mu(\alpha_i)\right) 
        & \text{ morphism }
        \\
        &= \pi\left(\mu(\alpha_0) \prod_{i = 1}^p \mu(u_i) \mu(\alpha_i)\right)
        & \text{ idempotent and } n_i \geq 1
        &= c \geq 0
    \end{align*}

    \textbf{Induction hypothesis.}
    The proof will essentially follow the same schema as the base case,
    with extra care needed to properly handle the summation of polynomials
    that arises when $k \geq 2$.

    We will use the fact that given $n_1, \dots, n_p \geq 0$,
    there exists a bijection between
    partitions of $q(\omega n_1, \dots, \omega n_p)$
    into $k+1$ words, and 
    triples composed of
    a position  $1 \leq \ell \leq p$,
    a ratio     $0 \leq b \leq n_\ell$,
    a remainder $0 \leq r < \omega \times \card{u_\ell}$,
    and a partition of
    $u_\ell^\omega\mid_{[-r:]} u_\ell^{\omega \times (n_\ell - b)}
    \alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{n_p} \alpha_{p}$
    into $k$ words,
    where we have used the notation 
    $v\mid_{[-r:]}$ to denote the word obtained from $v$ 
    by keeping only the last $r$ letters, in a sort of Pythonic syntax.
    That is, we look at where the first ``cut" is made.

    Now, given $(\ell, b, r)$ respecting the inequalities 
    mentioned above,
    one can define the \kl{polynomial functions}
    $s_{\ell,b,r} \colon \Nat^{\ell - 1} \to \Sigma^*$
    and
    $q_{\ell,b,r} \colon \Nat^{p - \ell + 1} \to \Sigma^*$
    so that
    $q(X_1, \dots, X_p) = s_{\ell, b, r}(X_1, \dots, X_{\ell-1}) 
    q_{\ell, b, r}(X_{\ell}, \dots, X_p)$
    under the assumption that $X_1 \geq b$:
    \begin{equation*}
        q_{\ell,b,r}(X_{1}, \dots, X_{p - \ell + 1})
        \defined
        u_\ell^\omega\mid_{[-r:]} u_\ell^{\omega \times (X_1 - b)}
        \alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{X_{p - \ell + 1}} \alpha_{p}
        \quad .
    \end{equation*}
    Similarly, let us define 
    $s_{\ell, b, r}(X_1, \dots, X_{\ell - 1})$
    as 
    $\alpha_0 u_1^{\omega X_1} \alpha_2 \dots \alpha_{\ell - 1} 
    u_\ell^{\omega b} u_{\ell}^\omega \mid_{[1:r]}$,
    Remark that $q_{\ell,b,r}$ has all of its indeterminates
    multiplied by $\omega$, so that the induction hypothesis 
    can be used.

    The key remark is that $\mu(s_{\ell,b,r}(X_1, \dots, X_{\ell - 1}))$ is a
    constant function from $\Nat^{\ell-1}$ to $M$, and that its value is also
    constant when $b$ varies among numbers greater than $0$. This holds with
    the exact same proof as the base case of this induction. Let us call
    $m_{\ell,r,1}$ and $m_{\ell,r,0}$ the two elements of the monoid that are
    obtained by $\mu(s_{\ell,b,r}(X_1, \dots, X_{\ell-1}))$ respectively when
    $b = 0$ and when $b > 0$. Finally, let us define $\pi_{\ell,r,0} \colon M^k
    \to \Nat$ via $\pi_{\ell,r,0}(m_2, \dots, m_{k+1}) \defined
    \pi(m_{\ell,r,0}, m_1, \dots, m_{k+1})$ and similarly for $\pi_{\ell,r,1}
    \colon M^k \to \Nat$.

    With all of this preliminary work done, we are now ready to conclude by
    simply unrolling the definition of 
    a function
    $f$ defined using a \kl{production function}
    $\pi \colon M^{k+1} \to \Nat$.
    In the following equations, we omit the precise
    range for the sum over $(\ell, b, r)$ to save space.
    Similarly, we will use 
    $q_{\ell,r}(X_1, \dots, X_{p-\ell+1})
    \defined q_{\ell,b,r}(X_1+b, X_2, \dots, X_{p - \ell + 1})$
    at the end of the computation.
    \begin{align*}
        &f(q(n_1, \dots, n_p))\\
        &\eqdef 
        \sum_{v_1 \dots v_{k+1} = q(n_1, \dots, n_p)}
        \pi(\mu(v_1), \dots, \mu(v_{k+1})) \\
        &= \sum_{(\ell, b, r)}
           \sum_{v_2 \dots v_{k+1} = q_{\ell,b,r}(n_\ell, \dots, n_p)}
           \pi(\mu(s_{\ell,b,r}(n_1, \dots, n_{\ell-1})), \mu(v_2), \dots, \mu(v_{k+1}))
        \\
        &= 
           \sum_{(\ell, 0, r)}
           \sum_{v_2 \dots v_{k+1} = q_{\ell,0,r}(n_\ell, \dots, n_p)}
           \pi(m_{\ell,r,0}, \mu(v_2), \dots, \mu(v_{k+1}))
           \\
        &+ 
           \sum_{(\ell, b > 0, r)}
           \sum_{v_2 \dots v_{k+1} = q_{\ell,b,r}(n_\ell, \dots, n_p)}
           \pi(m_{\ell,r,1}, \mu(v_2), \dots, \mu(v_{k+1}))
        \\
        &= 
           \sum_{(\ell, 0, r)}
           \sum_{v_2 \dots v_{k+1} = q_{\ell,0,r}(n_\ell, \dots, n_p)}
           \pi_{\ell,r,0}(\mu(v_2), \dots, \mu(v_{k+1}))
           \\
        &+ 
           \sum_{(\ell, b > 0, r)}
           \sum_{v_2 \dots v_{k+1} = q_{\ell,b,r}(n_\ell, \dots, n_p)}
           \pi_{\ell,r,1}(\mu(v_2), \dots, \mu(v_{k+1}))
        \\
        &= 
           \sum_{1 \leq \ell \leq p}
           \sum_{0 \leq r < \omega \times \card{u_\ell}}
           (\pi_{\ell,r,0})^\dagger (q_{\ell, 0, r}(n_\ell, \dots, n_p))
           \\
        &+ 
           \sum_{1 \leq \ell \leq p}
           \sum_{0 \leq r < \omega \times \card{u_\ell}}
           \sum_{0 < b < n_\ell}
           (\pi_{\ell,r,1})^{\dagger}(q_{\ell,r}(n_\ell - b, \dots, n_p))
    \end{align*}
    Now, using the induction hypothesis,
    we have \kl{natural binomial polynomials} $P_{\ell,r,0}$ and $P_{\ell,r,1}$
    satisfying 
    \begin{equation*}
       (\pi_{\ell,r,1})^{\dagger}(q_{\ell,r}(n_\ell - b, \dots, n_p))
       = P_{\ell,r,1}(\omega (n_\ell - b), \dots, \omega n_p)
       \quad 
       \forall n_\ell - b, \dots, n_p \geq 1
       \quad .
   \end{equation*}
   And similarly
    \begin{equation*}
        (\pi_{\ell,r,0})^{\dagger}(q_{\ell,r}(n_1, \dots, n_{\ell-1})
        = P_{\ell,r,0}(\omega n_1, \dots, \omega n_{\ell-1})
       \quad 
       \forall n_1, \dots, n_{\ell - 1} \geq 1
       \quad .
   \end{equation*}

   It is an easy and folklore fact that \kl{natural binomial polynomials} are
   stable under the operation $\Sigma_Y \colon Q(X_1,\dots,X_n,Y) \mapsto \sum_{0
   < b < Y} Q(X_1, \dots, X_n, b)$, and by pre-composition with 
   a multiplication by a constant $\omega \geq 1$.

   This allows us to conclude by defining 
   \begin{equation*}
       P(X_1, \dots, X_p) \defined
       \sum_{1 \leq \ell \leq p} \sum_{0 \leq r < \omega \card{u_\ell}}
       P_{l,r,0}(\omega X_{\ell}, \dots, \omega X_p) + \Sigma_Y P_{l,r,1}(\omega Y,
       \omega X_{\ell+1}, \dots, \omega X_p)
       \quad .
   \end{equation*}
   The latter being a \kl{natural binomial polynomial} satisfying
   the desired equality.
\end{proof}

Let us highlight that the multiplicative factor $\omega$ is necessary in
\cref{n-poly-combinatorics:lem}. Indeed, the function $f \colon \set{a}^* \to
\Nat$ defined as $0$ when the input is of odd length and $1$ when the input is
of even length is \kl{$\Nat$-polyregular}, but $f(a^X)$ is not a polynomial.

@@

\begin{corollary}
    \label{decide-rat-poly-npoly:cor}
    Let $P \in \Rat[\vec{X}]$ be a polynomial with \emph{rational}
    coefficients and let $\alpha$ be the smallest number in $\Nat_{\geq 1}$
    such that $\alpha P \in \Rel[\vec{X}]$. Then, the following are equivalent:
    \begin{enumerate}
        \item $P$ is a \kl{$\Nat$-rational polynomial},
        \item $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
        \item $\alpha P \in \CorrectPoly$,
            and for all $\vec{x} \in \Nat^{\vec{X}}$,
            $P(\vec{x})$ is a multiple of $\alpha$.
    \end{enumerate}
    In particular, the properties are decidable.
\end{corollary}
\begin{proof}
    If $P$ is a \kl{$\Nat$-rational polynomial}, then ...
    If $\alpha P \in \CorrectPoly$, then  it is an \kl{$\Nat$-rational polynomial}.
    But in particular, there exists a polyregular function $f$ that represents $\alpha P$
    with outputs in $\set{ a }$.
    Now, let us post-compose this function by the one that
    maps $a^{k \alpha + r}$ to $a^{k}$, which is a \kl{polyregular function}.
    We conclude that $P$ is \kl{represented} by an \kl{$\Nat$-polyregular function}.
\end{proof}
